---
title: "EXCHANGE Data Exploration"
output: html_document
date: "2022-10-07"
---

## Set-Up

```{r setup, include=FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE)

# load packages
library(tidyr)
library(broom)
library(ggplot2)
library(dplyr)
library(ggpubr)
library(viridis)
library(stringr)
library(janitor)
library(readxl)
```

```{r}
# build RData file for starting files - files imported in from csv downloads
# save(sample_collection_metadata, water_NPOC_TDN, site_metadata, water_spectral_indices, soil, water_spectral_indices_part2, file = "EC1_raw_data.RData")

load("EC1_raw_data.RData") 
# brings RData files into the environment!
```

## Data Preparation

```{r}
# simplify data frame to include coordinates and N/C metrics
water_sample_metadata <- sample_collection_metadata[, c(1:3, 7)]
water <- left_join(water_NPOC_TDN, water_sample_metadata, by = "kit_id")
water <- left_join(water, site_metadata, by = "kit_id")

# save a separate RData file for the site data!
water_sites <- sample_collection_metadata[,c(1:3)]
# save(water_sites, file = "water_collection_lat_long.RData")

# add a new variable to look at the major water body where samples were taken, not just the region
water <- water %>%
  mutate(water_body = case_when(
    state == "VA" | state == "MD" ~ "Chesapeake Bay",
    state == "DE" | state == "PA" | state == "NJ" ~ "Delaware Bay",
    kit_id == "K019" | kit_id == "K020" | kit_id == "K033" | kit_id == "K012" | kit_id == "K013" | kit_id == "K039" | kit_id == "K047" | kit_id == "K049" | state == "OH" ~ "Great Lakes"
  ))

# edit a few rows of the water type column
water$water_systemtype[11] <- "Tidal River/Estuary"
water$water_systemtype[23] <- "Tidal River/Lacestuary"
water$water_systemtype[43] <- "Tidal River"
water$water_systemtype[45] <- "Tidal Stream"

# set water type to factor variable
water$water_systemtype <- as.factor(water$water_systemtype)

# cut the dataset to the columns needed to make a map
water_map <- water[,c(3,5,6,9,10,11,15,21)]

# edit second CDOM dataset for better merging
colnames(water_spectral_indices_part2)[1] <- "kit_id"
water_spectral_indices_part2 <- water_spectral_indices_part2[,-2]

water_spectral_indices_part2$campaign <- "EC1"
water_spectral_indices_part2$transect_location <- "water"

# bring in the second set of CDOM data
water_spectral_indices_final <- rbind(water_spectral_indices, water_spectral_indices_part2)

# match up the spectral data with location of sites and DOC
water_CDOM <- left_join(water, water_spectral_indices_final, by = c("kit_id", "campaign", "transect_location"))

# pull out the kits without CDOM data to remove NAs - should just be 2 now with updated data
water_CDOM_final <- water_CDOM %>%
  filter(!is.na(SUVA254))

# remove additional columns for dataset ready for analysis
water_CDOM_final <- water_CDOM_final[,c(3,5,6,9:11,14,15,21:23,25,27:32)]

# make a list of the sites missing CDOM data
water_CDOM_missing <- water_CDOM %>%
  filter(is.na(SUVA254))

water_CDOM_missing <- water_CDOM_missing[,c(3,21)]

# remove kit 41 - can be replaced once that metadata is filled in
water_CDOM_final <- water_CDOM_final[-30,]
```

## Load In More Packages for Mapping

```{r, message=FALSE, warning=FALSE}
library(sf)
library(ggspatial)
# both of these packages integrate with ggplot for displaying spatial data

library(rnaturalearth)
library(rnaturalearthdata)
library(rnaturalearthhires)
# this package has important information about the US for drawing map boundaries
```

## Making Maps of Water DOC

```{r}
# pull info from rnaturalearth package to get world information
usa <- ne_states(country = "United States of America", returnclass = "sf")
gl <- ne_states(country = c("United States of America", "Canada"), returnclass = "sf")
# ne_states pulls geographic boundaries at the state level

#lakes <- rnaturalearth::ne_download(scale = 10, type = 'lakes', category = 'physical', returnclass = "sf")
# getting error message here to get more detail map attributes - address later!!
# https://www.naturalearthdata.com/downloads/10m-physical-vectors/ <- website for data downloads

# all water sampling locations plotted!
ggplot(data = usa) +
  geom_sf() +
  coord_sf(xlim = c(-95,-75), ylim = c(35,50)) + # set coordinate boundaries
  geom_point(data = water, mapping = aes(x = water_longitude, y = water_latitude, label = kit_id)) +
  geom_label(data = water, aes(x = water_longitude, y = water_latitude, label = kit_id))
# geom_label has to be fed the aesthetics to replace a point with the name of the point

ggplot(data = gl) +
  geom_sf()
```

```{r}
# divide dataframe into two by region for easier mapping
water$region <- as.factor(water$region)
CB_water <- water %>%
  filter(region == "Chesapeake Bay")
GL_water <- water %>%
  filter(region == "Great Lakes")

# zoomed in map for Chesapeake Region with NPOC values
ggplot(data = usa) +
  geom_sf() +
  coord_sf(xlim = c(-77.5,-74.5), ylim = c(37.2,40)) +
  geom_point(data = CB_water, mapping = aes(x = water_longitude, y = water_latitude, color = npoc_mgl)) + 
  # add points from another data frame
  theme_bw() +
  labs(x = "Longitude", y = "Latitude", color = "NPOC (mg/L)")

# zoomed in map for the Great Lakes
ggplot(data = gl) +
  geom_sf() +
  coord_sf(xlim = c(-92,-82), ylim = c(40,48)) +
  geom_point(data = GL_water, mapping = aes(x = water_longitude, y = water_latitude, color = npoc_mgl)) +
  theme_bw() +
  labs(x = "Longitude", y = "Latitude", color = "NPOC (mg/L)")
```

```{r}
# extra rough scatterplot to look if there are regional clusters in DOC concentrations for water
ggplot(na.omit(water_map), aes(x = water_latitude, y = npoc_mgl, color = water_body)) +
  geom_point() +
  theme_bw()
```

## Looking at Relationships between Different CDOM Spectral Indices

```{r}
# SUVA254 vs S350-400
ggplot(water_CDOM_final, aes(x = SUVA254, y = S350_400)) +
  geom_point() +
  theme_bw() +
  stat_cor()
```

```{r}
# S350-400 vs S275-295
ggplot(water_CDOM_final, aes(x = S275_295, y = S350_400)) +
  geom_point() +
  theme_bw() +
  stat_cor()
```

```{r}
# building a correlation plot!
# prepping data frame with just the variables for calculated correlations
water_corr_plot <- water_CDOM_final[,c(10:18)]

# load packages for visualization
library(ggcorrplot)

# compute the correlation matrix for NPOC and the calculated spectral indices
water_corr_matrix <- cor(water_corr_plot)

# calculate p-values for correlations
water_corr_pvalue <- cor_pmat(water_corr_plot)

# basic correlation plot with correlation coefficients
ggcorrplot(water_corr_matrix, hc.order = TRUE,
    type = "lower", lab = TRUE)

# make the correlation plot with significance values
ggcorrplot(water_corr_matrix, hc.order = TRUE,
    type = "lower", method = "circle", p.mat = water_corr_pvalue)
```

## Exploratory PCA (look at potentially redundant CDOM metrics)

```{r}
# https://www.statology.org/principal-components-analysis-in-r/
# set up data frame
water_pca <- water_CDOM_final[,c(10:18)]

# pull kit IDs to change the row names for labels on the PCA plot
kit_id <- water_CDOM_final[,1]

# change row names of data frame used for PCA
rownames(water_pca) <- kit_id
```

```{r}
# calculate the principal components
pc <- prcomp(water_pca, scale = TRUE)

# reverse signs of eigenvectors
pc$rotation <- -1*pc$rotation

# look at the principal components
pc$rotation

# reverse signs of each observation's principal components scores
pc$x <- -1*pc$x

# calculate variance represented by each of the principal components
pc$sdev^2 / sum(pc$sdev^2)
```

```{r}
# visualize the PCA
biplot(pc, scale = 0)

# use ggplot tool for PCA plotting to add outside metadata
library(ggfortify) #ggfortify allows ggplot to work with PCA data
# https://cran.r-project.org/web/packages/ggfortify/vignettes/plot_pca.html

autoplot(pc, data = water_CDOM_final, colour = 'water_body', size = 3, loadings = TRUE,
         loadings.colour = 'black', loadings.label = TRUE) +
  theme_bw()

# plot PCA again using water_systemtype
autoplot(pc, data = water_CDOM_final, colour = 'water_systemtype', size = 3, loadings = TRUE,
         loadings.colour = 'black', loadings.label = TRUE) +
  theme_bw()

```

```{r}
# plot SUVA on a map for regional trends (Chesapeake area specifically)
CDOM_CB <- water_CDOM_final %>%
  filter(region == "Chesapeake Bay")

ggplot(data = usa) +
  geom_sf() +
  coord_sf(xlim = c(-77.5,-74.5), ylim = c(37.2,40)) +
  geom_point(data = CDOM_CB, mapping = aes(x = water_longitude, y = water_latitude, color = SUVA254)) +
  theme_bw() +
  labs(x = "Longitude", y = "Latitude", color = "SUVA 254") +
  scale_color_viridis()
```

## Trying an NMDS to Look at Similar Water (in terms of CDOM metrics)

```{r}
# https://jonlefcheck.net/2012/10/24/nmds-tutorial-in-r/
# load necessary packages
library(vegan)

# run the NMDS use the data frame made for the PCA
site_NMDS = metaMDS(water_pca, k =2)

# Shepard plot to look at the stress
stressplot(site_NMDS)

# plot the NMDS
plot(site_NMDS)

# better plot with labels (run all lines together)
ordiplot(site_NMDS, type = "n")
orditorp(site_NMDS, display = "species")
orditorp(site_NMDS, display = "sites", cex = 1.25)
```

## Cluster Analysis (for a better visual for CDOM metrics)

```{r}
# http://www.sthda.com/english/wiki/beautiful-dendrogram-visualizations-in-r-5-must-known-methods-unsupervised-machine-learning
# https://www.r-bloggers.com/2021/04/cluster-analysis-in-r/

# compute Euclidean distance between samples
water_dist <- dist(scale(water_pca), method = "euclidean")

# hierarchical clustering based on the distance matrix
water_cluster <- hclust(water_dist, method = "complete")

# load package for dendrogram visualization
library(ggdendro)
library(dendextend) #used for more functionality with dendrograms

# draw the tree!
ggdendrogram(water_cluster)

# another variation of the tree to visualize the clusters directly (run both lines of code together)
plot(water_cluster, hang = -1) # hang = pulls all of the labels to the same location on the tree
rect.hclust(water_cluster, k =6)
```

```{r}
# pull data out of the dendrogram data class to be able to color the branches by a metadata variable
# https://towardsdatascience.com/custom-coloring-dendrogram-ends-in-r-f1fa45e5077a

# save dendrogram as an object
water_dendr <- as.dendrogram(water_cluster)

# extract dendrogram segement data to build own figure
dendr_data <- dendro_data(water_dendr)
dendr_segments <- dendr_data$segments

# pull out the segment ends and merge data with metadata for grouping by water body with colors
dendr_ends <- dendr_segments %>%
  filter(yend == 0) %>%
  left_join(dendr_data$labels, by = "x") %>%
  rename(kit_id = label) %>%
  left_join(water_CDOM_final, by = "kit_id")

# create vector of colors for the different regions
water_body_colors <- c("Chesapeake Bay" = "red" , "Delaware Bay" = "dark green", "Great Lakes" = "blue")

# plot the dendrogram
ggplot() +
 geom_segment(data = dendr_segments, aes(x=x, y=y, xend=xend, yend=yend)) +
 geom_segment(data = dendr_ends, aes(x=x, y=y.x, xend=xend, yend=yend, color = water_body)) + 
 scale_color_manual(values = water_body_colors) +
 scale_y_reverse() +
 coord_flip() + 
  theme_bw() + 
  ylab("Distance")
```

## Working with the Soil Data

```{r}
# remove all columns except metadata and TC data
soil_edit <- soil[,c(2,3,6)]

# prep water NPOC data for merging
NPOC_data <- water_map[,c(1,2,4:8)]
colnames(NPOC_data)[2] <- "water_npoc_mgl"

# merge two dataframes
water_soil_carbon <- left_join(NPOC_data, soil_edit, by = "kit_id")

# set three columns in dataset to be factor variables
water_soil_carbon$transect_location <- factor(water_soil_carbon$transect_location, levels = c("Wetland", "Transition", "Upland"))
water_soil_carbon$water_body <- as.factor(water_soil_carbon$water_body)
water_soil_carbon$region <- as.factor(water_soil_carbon$region)

# remove NA - should just be site 41 - step can be removed once metadata is replaced
# 57 and 60 have no soil samples
water_soil_carbon_noNA <- water_soil_carbon %>%
  filter(!is.na(region)) %>%
  filter(!is.na(tc_perc))
```

```{r}
# plot overall relationship between soil and water carbon
ggplot(na.omit(water_soil_carbon), aes(x = water_npoc_mgl, y = tc_perc, color = transect_location)) +
  geom_point() +
  facet_wrap(~transect_location, ncol = 2) +
  theme_bw() +
  stat_cor()
```

```{r}
# loop to do three different sets of plots (for transect location) faceted by location of sample

location_list <- c("Transition", "Upland", "Wetland") 
plot_list = list()
title_list <- c("Transition Soils", "Upland Soils", "Wetland Soils")

for (n in 1:3) {
  # will cycle through the loop one time for each site location
  filtered_data <- water_soil_carbon %>%
    filter(transect_location == location_list[n])
  
  # make a faceted plot
  plot <- ggplot(na.omit(filtered_data), aes(x = water_npoc_mgl, y = tc_perc, color = water_body)) +
  geom_point() +
  theme_bw() +
  labs(title = title_list[n]) + stat_cor()
  
  # add the plot to the plot_list for viewing
  plot_list[[n]] <- plot
}

print(plot_list)
```

```{r}
# new plots looking at water_systemtype

second_plot_list = list()

for (n in 1:3) {
  # will cycle through the loop one time for each site location
  filtered_data <- water_soil_carbon %>%
    filter(transect_location == location_list[n])
  
  # make a faceted plot
  plot <- ggplot(na.omit(filtered_data), aes(x = water_npoc_mgl, y = tc_perc, color = water_systemtype)) +
  geom_point() +
  facet_wrap(~water_systemtype, ncol = 4) +
  theme_bw() +
  labs(title = title_list[n]) + stat_cor()
  
  # add the plot to the plot_list for viewing
second_plot_list[[n]] <- plot
}

print(second_plot_list)
```

```{r}
# boxplot of the soil carbon percentages by transect location
ggplot(water_soil_carbon_noNA, aes(x=water_body, y = tc_perc, fill = transect_location)) +
  geom_boxplot() +
  theme_bw()

# separate out the water system type
ggplot(water_soil_carbon_noNA, aes(x=water_body, y = tc_perc, fill = transect_location)) +
  geom_boxplot() +
  theme_bw() +
  facet_wrap(~ water_systemtype)
```

## Thinking through Set of Soil Extractions

```{r}
# bring in metadata about soil samples
soil_availability <- readxl::read_excel("~/EC1 Soil Jar Sieving Metadata.xlsx")

# make spreadsheet column names tidy
soil_availability <- clean_names(soil_availability)

# remove excess columns
soil_availability_reduce <- soil_availability[,-c(2:5,19:32)]

# pull out the transect location from sample name and remove it from the sample name
soil_availability_reduce <- soil_availability_reduce %>%
  mutate(transect_location = gsub(".*_", "", sample),
         kit_id = gsub("_.*", "", sample)) %>%
  select(-sample) %>%
  relocate(transect_location, .before = sample_weight_from_tare_g) %>%
  relocate(kit_id, .before = transect_location) %>%
  mutate(transect_location = str_to_title(transect_location))

# add column for whether or not a kit has sediment
soil_availability_reduce$sediment_available <- ifelse(soil_availability_reduce$transect_location == "Sediment",
                                                      1,
                                                      0)

# summary data frame from the availability spreadsheet
available_soils_sum <- soil_availability_reduce %>%
  group_by(kit_id) %>%
  summarize(TAI_gradient = n(),
            sediment = sum(sediment_available))

# figure out if UP -> W is available
available_soils_sum <- available_soils_sum %>%
  mutate(UP_T_W = TAI_gradient - sediment) %>%
  select(-TAI_gradient)
```

```{r}
# prep metadata data frames!!
# group the soil samples by Kit ID and then count how many per kit
site_summary <- water_soil_carbon_noNA %>%
  group_by(kit_id, water_npoc_mgl, water_latitude, water_longitude, region, water_systemtype, water_body) %>%
  summarize(soil_samples = n())

# bring salinity data in!
water_quality <- read.csv("~/EC1_WaterQuality_L0B_20220509.csv")
site_summary_WQ <- left_join(site_summary, water_quality, by = "kit_id")

# merge soil availability with the metadata
all_available_kits <- left_join(available_soils_sum, site_summary_WQ)
```

```{r}
# FILTERING TO TARGET SITES
filter_available_kits <- all_available_kits %>%
  filter(!is.na(water_npoc_mgl)) %>% # remove sites without water information
  filter(UP_T_W == 3) %>% # remove sites that do not have the entire transect present
  select(-transect_location) # remove this column for better merging

# bring in excel sheet with soil amounts classified
soil_availability_class <- readxl::read_excel("~/EC1_WSOM_Feasibility.xlsx")

# add a column to indicate whether a kit can absolutely happen or not
soil_availability_class$possible_at_all <- ifelse(soil_availability_class$status == "nope",
                                            1,
                                            0)

# add column for indication of hard to do
soil_availability_class$hard_to_do <- ifelse(soil_availability_class$status == "not worth it",
                                       1,
                                       0)

# add column to indicate ones that we would need a good argument for
soil_availability_class$strong_case <- ifelse(soil_availability_class$status == "unlikely",
                                        1,
                                        0)

# add column to assign points to the rest of the categories
soil_availability_class$points <- case_when(
  soil_availability_class$status == "okay" ~ 1,
  soil_availability_class$status == "more likely" ~ 2,
  soil_availability_class$status == "best case scenario" ~ 3,
  soil_availability_class$status == "already analyzed" ~ 5
)
soil_availability_class$points <- ifelse(is.na(soil_availability_class$points),
                                   0,
                                   soil_availability_class$points)

# summary data sheet
soil_class_sum <- soil_availability_class %>%
  group_by(kit_id) %>%
  summarize(total_points = sum(points),
            total_strong_case = sum(strong_case),
            total_hard = sum(hard_to_do),
            total_nope = sum(possible_at_all))

# create the final score for each of the kits
soil_class_sum$final_score <- soil_class_sum$total_points

for (a in 1:nrow(soil_class_sum)) {
  
  if (soil_class_sum$total_nope[a] != 0) {
    soil_class_sum$final_score[a] <- 0
  }
  else {
    if (soil_class_sum$total_hard[a] != 0) {
      soil_class_sum$final_score[a] <- 1
    }
    else {
      if (soil_class_sum$total_strong_case[a] != 0) {
        soil_class_sum$final_score[a] <- 2
      }
    }
  }
}

# merge these scores with the filtered kits metadata
target_sites <- left_join(soil_class_sum, filter_available_kits, by = "kit_id")

# remove unnecessary columns
target_sites <- target_sites[,-c(8,15,16,21:24)]

# make final score a factor variable for easier mapping
target_sites$final_score <- as.factor(target_sites$final_score)
```

```{r}
# maps of kits left that fit the criteria

# map the Delaware Bay information
ggplot(data = usa) +
  geom_sf() +
  coord_sf(xlim = c(-75.75,-74.75), ylim = c(38.9,40)) +
  geom_point(data = filter(target_sites, water_body == "Delaware Bay"), mapping = aes(x = water_longitude, y = water_latitude, color = final_score), size = 2.5) +
  theme_bw() +
  labs(x = "Longitude", y = "Latitude", color = "Feasibility Score")

# map the Chesapeake Bay
ggplot(data = usa) +
  geom_sf() +
  coord_sf(xlim = c(-77.25,-75.5), ylim = c(37.2,39)) +
  geom_point(data = filter(target_sites, water_body == "Chesapeake Bay"), mapping = aes(x = water_longitude, y = water_latitude, color = final_score), size = 2.5) +
  theme_bw() +
  labs(x = "Longitude", y = "Latitude", color = "Feasibility Score")

# map the Great Lakes
ggplot(data = gl) +
  geom_sf() +
  coord_sf(xlim = c(-92,-82), ylim = c(41.5,47)) +
  geom_point(data = filter(target_sites, water_body == "Great Lakes"), mapping = aes(x = water_longitude, y = water_latitude, color = final_score), size = 2.5) +
  theme_bw() +
  labs(x = "Longitude", y = "Latitude", color = "Feasibility Score")
```

``` {r}
# map target sites to get a feel for spatial layout
# chesapeake region
ggplot(data = usa) +
  geom_sf() +
  coord_sf(xlim = c(-77.5,-74.5), ylim = c(37.2,40)) +
  geom_point(data = target_sites, mapping = aes(x = water_longitude, y = water_latitude, color = water_systemtype)) +
  theme_bw() +
  labs(x = "Longitude", y = "Latitude")

# chesapeake with a salinity gradient
ggplot(data = usa) +
  geom_sf() +
  coord_sf(xlim = c(-77.5,-74.5), ylim = c(37.2,40)) +
  geom_point(data = target_sites, mapping = aes(x = water_longitude, y = water_latitude, color = sal_psu)) +
  theme_bw() +
  labs(x = "Longitude", y = "Latitude")

# GL region
ggplot(data = gl) +
  geom_sf() +
  coord_sf(xlim = c(-92,-82), ylim = c(40,48)) +
  geom_point(data = target_sites, mapping = aes(x = water_longitude, y = water_latitude, color = water_systemtype)) +
  theme_bw() +
  labs(x = "Longitude", y = "Latitude")
```

```{r}
# old vs new selected sites
final_extract_sites <- target_sites %>%
  filter(kit_id == "K019" |
           kit_id == "K013" |
           kit_id == "K010" |
           kit_id == "K050" |
           kit_id == "K061" |
           kit_id == "K048" |
           kit_id == "K026" |
           kit_id == "K018" |
           kit_id == "K035" |
           kit_id == "K025")

original_extract_sites <- target_sites %>%
    filter(kit_id == "K019" |
           kit_id == "K013" |
           kit_id == "K010" |
           kit_id == "K039" |
           kit_id == "K056" |
           kit_id == "K048" |
           kit_id == "K040" |
           kit_id == "K015" |
           kit_id == "K035" |
           kit_id == "K044")

# plot original sites!!
ggplot(data = gl) +
  geom_sf() +
  coord_sf(xlim = c(-92,-75), ylim = c(37,47)) +
  geom_point(data = original_extract_sites, mapping = aes(x = water_longitude, y = water_latitude, color = sal_psu), size = 2) +
  theme_bw() +
  labs(x = "Longitude", y = "Latitude")

ggplot(data = usa) +
  geom_sf() +
  coord_sf(xlim = c(-77.5,-74.5), ylim = c(37.2,40)) +
  geom_point(data = original_extract_sites, mapping = aes(x = water_longitude, y = water_latitude, color = sal_psu), size = 2) +
  theme_bw() +
  labs(x = "Longitude", y = "Latitude")

# plot final sites!!
ggplot(data = gl) +
  geom_sf() +
  coord_sf(xlim = c(-92,-75), ylim = c(37,47)) +
  geom_point(data = final_extract_sites, mapping = aes(x = water_longitude, y = water_latitude, color = sal_psu), size = 2) +
  theme_bw() +
  labs(x = "Longitude", y = "Latitude")

ggplot(data = usa) +
  geom_sf() +
  coord_sf(xlim = c(-77.5,-74.5), ylim = c(37.2,40)) +
  geom_point(data = final_extract_sites, mapping = aes(x = water_longitude, y = water_latitude, color = sal_psu), size = 2) +
  theme_bw() +
  labs(x = "Longitude", y = "Latitude")
```

```{r}
# bring in soils metadata to look at gradients in characteristics
all_soil_info <- readRDS("soils_data_merged_withmeta (4).rds")

# cut soil information down to the sites used
soil_info_sites <- all_soil_info %>%
  filter(kit_id %in% final_extract_sites$kit_id)

# get lat/long info for the WSOM extracts
sites_lat_long <-sample_collection_metadata %>%
  filter(kit_id %in% final_extract_sites$kit_id) %>%
  select(kit_id, contains("latitude")|contains("longitude")) %>%
  select(-contains("sediment")) %>%
  pivot_longer(-kit_id, 
               names_to = c("transect_location", ".value"),
               names_pattern = "(.+)_(.+)") %>%
  mutate(transect_location = str_to_title(transect_location))

# create data frame with lat/long and soil characteristics info for the extract sites
extract_sites_soil_info <- soil_info_sites %>%
  select(-contains("flag"), -date_run, -contains("collect")) %>%
  filter(transect_location != "Sediment") %>%
  full_join(sites_lat_long, by = c("kit_id", "transect_location"))
  
```

```{r}
# look at all sites in the Chesapeake that could be extracted
target_sites_soils_info <- all_soil_info %>%
  filter(state == "VA" | state == "MD") %>%
  filter(kit_id %in% target_sites$kit_id) %>%
  filter(transect_location != "Sediment")

target_sites_lat_long <- sample_collection_metadata %>%
  filter(kit_id %in% target_sites_soils_info$kit_id) %>%
  select(kit_id, contains("latitude")|contains("longitude")) %>%
  select(-contains("sediment")) %>%
  pivot_longer(-kit_id, 
               names_to = c("transect_location", ".value"),
               names_pattern = "(.+)_(.+)") %>%
  mutate(transect_location = str_to_title(transect_location))

target_sites_soils_info <- target_sites_soils_info%>%
  select(-contains("flag"), -date_run, -contains("collect")) %>%
  full_join(target_sites_lat_long, by = c("kit_id", "transect_location"))

# bring in water data too
target_sites_water <- target_sites_soils_info %>%
  filter(transect_location == "Water") %>%
  select(kit_id, transect_location, latitude, longitude) %>%
  left_join(select(target_sites, kit_id, water_npoc_mgl, sal_psu, ph, orp_mv, alk_mgl_caco3))

# bind soils and water info together
target_sites_soils_info <- target_sites_soils_info %>%
  filter(transect_location != "Water")

target_sites_soils_water <- bind_rows(target_sites_soils_info, target_sites_water) %>%
  group_by(kit_id) %>%
  mutate(already_extracted = ifelse(kit_id %in% final_extract_sites$kit_id,
                                    "YES",
                                    "NO"))

# use newly created dataset to look at range in specific conductance and pH
target_sites_soils_water %>%
  filter(transect_location != "Water") %>%
  ggplot(aes(x = factor(transect_location, levels = c("Upland", "Transition", "Wetland")), y = specific_conductance_us_cm, color = kit_id, shape = already_extracted)) +
  geom_point() +
  labs(x = "Transect Location")

target_sites_soils_water %>%
  ggplot(aes(x = factor(transect_location, levels = c("Upland", "Transition", "Wetland", "Water")), y = ph, color = kit_id, shape = already_extracted)) +
  geom_point() +
  labs(x = "Transect Location")

```

```{r}
# zoom in on sites already extracted
# look at values of specific conductance by kit
target_sites_soils_water %>%
  filter(already_extracted == "YES") %>%
  filter(transect_location != "Water") %>%
  ggplot(aes(x = factor(transect_location, levels = c("Upland", "Transition", "Wetland")), y = specific_conductance_us_cm, color = kit_id)) +
  geom_point() +
  labs(x = "Transect Location")

# look at values of pH by kit
target_sites_soils_water %>%
  filter(already_extracted == "YES") %>%
  ggplot(aes(x = factor(transect_location, levels = c("Upland", "Transition", "Wetland", "Water")), y = ph, color = kit_id)) +
  geom_point() +
  labs(x = "Transect Location")

```


