---
title: "WSOM Extracts Analysis"
output: html_document
date: "2022-11-24"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load Packages and Data

```{r, include=FALSE, warning = FALSE, message = FALSE}
# required packages
packages_list <- c("tidyverse", "broom", "ggpubr", "sf", "ggspatial",
                   "rnaturalearth", "rnaturalearthdata",
                   "rnaturalearthhires", "rstatix", "ggcorrplot",
                   "GGally")

lapply(packages_list, library, character.only = TRUE)

# set data files location
data_location = paste0(getwd(), "/R Data Files")

# RDS files - master water/soil data, sample site coordinates, wsom concentrations
water_soil <- read_rds(paste0(data_location, "/EC1_water_soil_data.rds"))
site_coordinates <- read_rds(paste0(data_location, 
                                    "/sample_collection_coordinates.rds"))
wsom_conc <- read_rds(paste0(data_location, "/wsom_conc.rds"))

# change region name for Chesapeake to Mid-Atlantic (for figures)
water_soil <- water_soil %>%
  mutate(region = ifelse(region == "Chesapeake Bay", "Mid-Atlantic", "Great Lakes"))
water_soil$region <- as.factor(water_soil$region)
```

## Set-Up

```{r, echo=FALSE, include=FALSE, message=FALSE, warning=FALSE, eval=FALSE}
# set up fonts for figures
#install.packages("showtext")
#library(showtext)

# check available fonts to import
#font_files()

# pull in cambria
#font_add("Cambria", regular = "C:/Windows/Fonts/cambria.ttc", italic = "C:/Windows/Fonts/cambriai.ttf", bold = "C:/Windows/Fonts/cambriab.ttf")
```

```{r}
# set ggplot theme - remove dark background
theme_set(theme_bw() +
            theme(text = element_text(family = "serif"),
                  axis.title = element_text(size = 20),
                  axis.text = element_text(size = 18),
                  legend.text = element_text(size = 18),
                  legend.title = element_text(size = 20)))

# create color palette
palette <- c("#056009","#0385B5","#021677","#ABABD9","#27220F","#8E7941",
             "#DCCD6A","#8F8D90")

# destination for figure output
output_directory <- "./Saved Figures"
```

## Map Figures

```{r}
# pull info from rnaturalearth
usa <- ne_states(country = "United States of America", returnclass = "sf")
gl <- ne_states(country = c("United States of America", "Canada"), 
                returnclass = "sf")

# filter coordinates to water sites only for mapping
water_gps <- site_coordinates %>%
  filter(transect_location == "Water")

# plot all EXCHANGE water sites
ggplot(data = gl) +
  geom_sf(fill = "#EEEEEE") +
  coord_sf(xlim = c(-92,-74), ylim = c(37,48)) +
  geom_point(data = water_gps, mapping = aes(x = longitude, y = latitude)) +
  labs(x = "Longitude", y = "Latitude") +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
  # last line removes grid lines from the map

# crop to Chesapeake
ggplot(data = gl) +
  geom_sf(fill = "#EEEEEE") +
  coord_sf(xlim = c(-78,-74), ylim = c(37,40.5)) +
  geom_point(data = water_gps, mapping = aes(x = longitude, y = latitude)) +
  labs(x = "Longitude", y = "Latitude") +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
```

```{r}
# make a list of my sites for filtering the data
extract_sites <- list("K010", "K013", "K018", "K019", "K025", "K026", "K035",
                      "K048", "K050", "K061")

# filter site coordinates (and remove sediment points)
extracts_gps <- site_coordinates %>%
  filter(kit_id %in% extract_sites & transect_location != "Sediment")

# just plot water for general site locations
ggplot(data = gl) +
  geom_sf(fill = "#EEEEEE") +
  coord_sf(xlim = c(-92,-74), ylim = c(37,47)) +
  geom_point(data = filter(extracts_gps, transect_location == "Water"), 
             mapping = aes(x = longitude, y = latitude),
             color = "#021677", size = 2.5) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  labs(x = "", y = "")

# export map for presentation
ggsave("site_map.png", plot = last_plot(), path = output_directory)

# zoom in on chesapeake and plot the full transect
ggplot(data = gl) +
  geom_sf(fill = "#EEEEEE") +
  coord_sf(xlim = c(-78,-75), ylim = c(37,40)) +
  geom_point(data = extracts_gps, mapping = aes(x = longitude, y = latitude),
             color = "#021677", size = 2.5) +
  scale_x_continuous(breaks = seq(-78,-75, by = 1)) +
  labs(x = "", y = "") +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())

# export map
ggsave("chesapeake_site_map.png", plot = last_plot(), path = output_directory)
```

## Patterns in WSOC and WSON

```{r}
# filter water/soil characteristics and merge with WSOM info
wsom_extracts <- water_soil %>%
  filter(kit_id %in% wsom_conc$kit_id & transect_location != "Sediment") %>%
  left_join(wsom_conc, by = c("campaign", "kit_id", "transect_location"))

# relevel transect_location variable
wsom_extracts$transect_location <- factor(wsom_extracts$transect_location,
                                          levels = c("Upland", "Transition",
                                                     "Wetland"))
```

```{r}
# density plot to look at WSOC and WSON value distributions
ggplot(wsom_extracts, aes(x = wsoc_conc)) +
  geom_density()

ggplot(wsom_extracts, aes(x = wson_conc)) +
  geom_density()

# both are non-normally distributed (gamma distribution)
# move forward using non-parametric statistical tests
```

```{r, warning = FALSE}
# boxplot for WSOC by transect location
ggplot(wsom_extracts) +
  geom_boxplot(aes(x = transect_location, y = wsoc_conc, fill = transect_location),
               show.legend = FALSE) +
  scale_fill_manual(values = palette) +
  labs(x = "Gradient Location", y = "WSOC Concentration (mg/g soil)")

ggsave("wsoc_location_boxplot.png", path = output_directory)

# non-parametric ANOVA
kt_wsoc <- kruskal.test(wsoc_conc ~ transect_location, data = wsom_extracts)
dunn_test(wsom_extracts, wsoc_conc ~ transect_location)

# boxplot for WSOC per g moist soil by transect location
# testing to see if true field conditions shows a difference
ggplot(wsom_extracts) +
  geom_boxplot(aes(x = transect_location, y = wsoc_conc_field, fill = transect_location),
               show.legend = FALSE) +
  scale_fill_manual(values = palette) +
  labs(x = "Gradient Location", y = "WSOC Concentration (mg/g field soil)")

# non-parametric ANOVA for field moist concentrations
kt_wsoc <- kruskal.test(wsoc_conc_field ~ transect_location, data = wsom_extracts)
dunn_test(wsom_extracts, wsoc_conc_field ~ transect_location)

# still not a statistically significant difference - almost uplands vs the other two
# one value missing because of missing GWC but definitely outlier site
```

```{r}
# test line graph to look at changes across the transect in WSOC
a <- ggplot(wsom_extracts, aes(x = transect_location, y = wsoc_conc, color = kit_id,
                          group = kit_id)) +
  geom_point() +
  geom_line() +
  facet_wrap(~water_body)

b <- ggplot(wsom_extracts, aes(x = transect_location, y = tc_perc, color = kit_id,
                               group = kit_id)) +
  geom_point() +
  geom_line() +
  facet_wrap(~water_body)

ggarrange(a,b, ncol = 1,
          common.legend = TRUE, legend = "right")
```

```{r}
# boxplot for WSOC regionally
ggplot(wsom_extracts) +
  geom_boxplot(aes(x = region, y = wsoc_conc, fill = transect_location)) +
  scale_fill_manual(values = palette) +
  labs(x = "Region", y = "WSOC Concentration (mg/g soil)", fill = "Transect Location")

# non-parametric ANOVA
kt_wsoc_region <- kruskal.test(wsoc_conc ~ region, data = wsom_extracts)
dunn_test(wsom_extracts, wsoc_conc ~ region)
```

```{r}
# setting up data for friedman test - needs to be a "blocked" design
wsoc_friedman <- wsom_extracts %>%
  select(c(wsoc_conc, transect_location, region))

# friedman test - equivalent of 2-way ANOVA
friedman.test(wsoc_conc ~ transect_location | region, data = wsoc_friedman)

# data is not structured correctly for friedman test - not equivalent to a two-way ANOVA - data needs to be from a nonreplicated block design, meaning one observation per intersection of the two variables
```

```{r}
# trying multiple Mann-Whitney U tests to test differences between combos of region and transect location
upland_gl <- wsom_extracts %>%
  filter(region == "Great Lakes" & transect_location == "Upland") %>%
  select(wsoc_conc)
colnames(upland_gl) <- "up_gl_wsoc"

transition_gl <- wsom_extracts %>%
  filter(region == "Great Lakes" & transect_location == "Transition") %>%
  select(wsoc_conc)
colnames(transition_gl) <- "t_gl_wsoc"

wetland_gl <- wsom_extracts %>%
  filter(region == "Great Lakes" & transect_location == "Wetland") %>%
  select(wsoc_conc)
colnames(wetland_gl) <- "w_gl_wsoc"

upland_ma <- wsom_extracts %>%
  filter(region == "Mid-Atlantic" & transect_location == "Upland") %>%
  select(wsoc_conc)
colnames(upland_ma) <- "up_ma_wsoc"

transition_ma <- wsom_extracts %>%
  filter(region == "Mid-Atlantic" & transect_location == "Transition") %>%
  select(wsoc_conc)
colnames(transition_ma) <- "t_ma_wsoc"

wetland_ma <- wsom_extracts %>%
  filter(region == "Mid-Atlantic" & transect_location == "Wetland") %>%
  select(wsoc_conc)
colnames(wetland_ma) <- "w_ma_wsoc"

grouped_wsoc_conc <- upland_gl
grouped_wsoc_conc$t_gl_wsoc <- transition_gl$t_gl_wsoc
grouped_wsoc_conc$w_gl_wsoc <- wetland_gl$w_gl_wsoc
grouped_wsoc_conc$up_ma_wsoc <- upland_ma$up_ma_wsoc
```

```{r}
# boxplot for tc percent
ggplot(data = wsom_extracts, aes(x = transect_location, y = tc_perc, fill = transect_location)) +
  geom_boxplot(show.legend = FALSE) +
  scale_fill_manual(values=palette) +
  labs(x = "Gradient Location", y = "Total Carbon %")

# kruskal-wallis test
kt_tc <- kruskal.test(tc_perc ~ transect_location, wsom_extracts)
dunn_test(wsom_extracts, tc_perc ~ transect_location)

# boxplot and kruskal-wallis test for tc across all sites (not just extract sites)
ggplot(filter(water_soil, transect_location == "Upland"|transect_location == "Transition" | transect_location == "Wetland"), 
       aes(x = transect_location, y = tc_perc, fill = transect_location)) +
  geom_boxplot(show.legend = FALSE) +
  scale_fill_manual(values = palette) +
  labs(x = "Transect Location", y = "Total Carbon (%)")

kt_tc_all <- kruskal.test(tc_perc ~ transect_location, water_soil)
dunn_test(water_soil, tc_perc ~ transect_location)
```

```{r}
# boxplot for WSON
ggplot(wsom_extracts) +
  geom_boxplot(aes(x = transect_location, y = wson_conc, fill = transect_location),
               show.legend = FALSE) +
  scale_fill_manual(values = palette) +
  labs(x = "Location", y = "WSON Concentration (mg/g soil)")
```

## WSOM and Other Soil Characteristics

### Correlation Plot for WSOC

```{r}
# pull variables
wsoc_corr <- wsom_extracts[,c(16:18,20,29)]

# make correlation matrix
wsoc_corr_matrix <- cor(wsoc_corr, use = "complete.obs")
wsoc_corr_pvalue <- cor_pmat(wsoc_corr)

# plot correlation matrix
ggcorrplot(wsoc_corr_matrix, hc.order = TRUE,
    type = "lower", lab = TRUE)
```

```{r, warning = FALSE, message = FALSE}
# matrix of scatterplots
# set the grouping variable for graphing
groups <- wsom_extracts[,15]

# plot with base R
pairs(wsoc_corr,
      labels = c("Bulk Density (g/cm3)", "LOI (%)", "GWC (%)", "Total Carbon %", "pH",
                 "WSOC (mg/g soil)"),
      pch = 21,
      bg = palette[c(1:3)][groups],
      col = palette[c(1:3)][groups],
      upper.panel = NULL)

# plot with ggplot
ggpairs(wsoc_corr, ggplot2::aes(color = groups),
        #lower = list(continuous = "smooth"), # adds lm line to the scatterplots
        columnLabels = c("Bulk Density (g/cm3)", "LOI (%)", "GWC (%)",
                         "Total Carbon (%)", "WSOC (mg/g soil)")) +
  ggplot2::scale_fill_manual(values = palette) +
  ggplot2::scale_color_manual(values = palette)
```

### Relationship between Total Carbon % and WSOC Concentration

```{r}
# soil TC vs WSOC plot
ggplot(wsom_extracts, aes(x = tc_perc, y = wsoc_conc)) +
  geom_smooth(formula = y~x, method = "lm", se = FALSE, color = palette[8]) +
  labs(x = "Soil Total Carbon %", y = "WSOC Concentration (mg/g soil)", color = "Gradient Location") +
  scale_color_manual(values = palette) +  
  geom_point(aes(color = transect_location), size = 2.5) +
  stat_regline_equation(aes(label = ..rr.label..), size = 5)

# stat_regline_equation is part of the ggpubr package
# can also put the full equation (label = ..eq.label..), AIC, BIC...

# save plot
ggsave("wsoc_tc_scatter.png", plot = last_plot(), path = output_directory,
       width = 7, height = 5)
```

```{r}
# fit a linear model (w/ transect location)
wsoc_tc <- lm(wsoc_conc ~ tc_perc + transect_location, data = wsom_extracts)

# look at full output
summary(wsoc_tc)

# model w/o location
lm(wsoc_conc ~ tc_perc, data = wsom_extracts) %>%
  summary()

lm(wsoc_conc~ tc_perc + transect_location + region, wsom_extracts) %>%
  summary()
```

### Relationship between WSOC and LOI

```{r}
ggplot(wsom_extracts, aes(x = loi_perc, y = wsoc_conc)) +
  geom_smooth(formula = y~x, method = "lm", se = FALSE, color = palette[8]) +
  labs(x = "Loss on Ignition (%)", y = "WSOC Concentration (mg/g soil)", color = "Gradient Location") +
  scale_color_manual(values = palette) +  
  geom_point(aes(color = transect_location), size = 2.5) +
  stat_regline_equation(aes(label = ..rr.label..), size = 5)

ggsave("wsoc_loi.png", plot = last_plot(), path = output_directory,
       width = 7, height = 5)
```

```{r}
lm(wsoc_conc ~ loi_perc, data = wsom_extracts) %>%
  summary()

lm(wsoc_conc ~ loi_perc + transect_location, data = wsom_extracts) %>%
  summary()

lm(wsoc_conc ~ loi_perc + transect_location + region, data = wsom_extracts) %>%
  summary()
```

### Relationship between Total Nitrogen % and WSON

```{r}
ggplot(wsom_extracts, aes(x = tn_perc, y = wson_conc)) +
  geom_smooth(formula = y~x, method = "lm", se = FALSE, color = palette[8]) +
  labs(x = "Soil Total Nitrogen %", y = "WSON Concentration (mg/g soil)", color = "Location") +
  scale_color_manual(values = palette) +  
  geom_point(aes(color = transect_location), size = 2.5) +
  stat_regline_equation(aes(label = ..rr.label..))
```

## WSOM and Surface Water DOM

### Relationship between Soil WSOM at Different Transect Location and Aquatic DOM

```{r}
ggplot(wsom_extracts, aes(x=wsoc_conc, y=water_npoc_mgl, color = transect_location)) +
  geom_point()+
  scale_color_manual(values = palette) +
  facet_wrap(~transect_location)

ggplot(wsom_extracts, aes(x=tc_perc, y = water_npoc_mgl, color = transect_location)) +
  geom_point() +
  scale_color_manual(values = palette) +
  facet_wrap(~transect_location)
```
