---
title: "WSOM and GHG Exploration"
output: html_document
date: "2023-03-06"
---

This file reads in spectral data from WSOM extracts and GHG data from lab experiments.
The goal is to look at preliminary relationships between the composition of WSOM across
the transect locations in comparison with metrics related to microbial processing (oxygen consumption, GHG production).

## Set-Up

```{r setup, include=FALSE}
# load packages
require(pacman)
pacman::p_load(tidyverse,
               janitor,
               broom,
               ggpubr, # to use ggarrange for plots
               ggfortify) # for plotting PCA

# load necessary data
ghg <- read_csv("./Original Data Files/EC1_GHG_data_PR.csv")
wsom_cdom <- read_csv("./Original Data Files/EC1_WSOC_Extracts_Nov2022_SpectralIndices.csv")
```

```{r}
# set ggplot theme
theme_set(theme_bw())

# set palette
palette <- c("#056009","#0385B5","#021677","#ABABD9","#27220F","#8E7941",
             "#DCCD6A","#8F8D90")

```


## Data Frame Edits

```{r edit-wsom-df}
# pull kit_id and transect location out of cdom sample_id
wsom_cdom$kit_id <- NA
wsom_cdom$transect_location <- NA

for (a in 1:nrow(wsom_cdom)) {
  
  # pull out the second piece of the sample ID split apart at the underscore and use as the plot
  wsom_cdom$kit_id[a] <- strsplit(wsom_cdom$Sample_ID, " ")[[a]][1]

  # pull out the fourth piece of that string and use as the timepoint
  wsom_cdom$transect_location[a] <- strsplit(wsom_cdom$Sample_ID, " ")[[a]][2]
  
}

# mutate transect location abbreviations to match ghg data frame
wsom_cdom_edit <- wsom_cdom %>%
  mutate(transect_location = case_when(
    transect_location == "T" ~ "Transition",
    transect_location == "W" ~ "Wetland",
    transect_location == "UP" ~ "Upland"
  )) %>%
  # remove tea standards
  filter(!is.na(transect_location))

# create unique list of kit IDs and transect locations
unique_extracts <- wsom_cdom_edit %>%
  select(kit_id, transect_location) %>%
  unique()

```

```{r edit-ghg-df}
# filter ghg data to just kits that also have extract data
ghg_filter <- ghg %>%
  filter(kit_id %in% unique_extracts$kit_id) %>%
  filter(transect_location %in% unique_extracts$transect_location)

# trim wsom data frame to match ghg and merge the two together
ghg_wsom_extracts <- wsom_cdom_edit %>%
  filter(kit_id %in% ghg_filter$kit_id) %>%
  # remove the replicate rows - could edit to take the average of these rows to move forward
  filter(!grepl("Rep", Sample_Description)) %>%
  # select specific indices
  select(kit_id, transect_location, Sample_ID:SUVA254, S350_400, FI:FRESH) %>%
  full_join(ghg_filter, by = c("kit_id", "transect_location")) %>%
  # reorder transect location
  mutate(transect_location = factor(transect_location, levels = c("Upland", "Transition", "Wetland")))

# make long data frame for the indices to be able to facet the exploratory plots
ghg_wsom_long <- ghg_wsom_extracts %>%
  pivot_longer(cols = c(SUVA254:FRESH), names_to = "spectral_index", values_to = "index_value")

```

## Exploratory Plots

### Dissolved Oxygen Consumption

First thoughts - higher rates of DO consumption should correlate with higher levels of microbial processing.
With more microbial processing, SUVA254 should decline and inversely S350-400 would increase.
FRESH, FI, and HIX should also increase?!?! The question is are these patterns related to transect location as well?

```{r do-vs-indices-scatterplot}
# SUVA254
ggplot(ghg_wsom_extracts, aes(x = delta_do_hr, y = SUVA254, color = transect_location)) +
  geom_point() +
  # add limit because of some outliers
  xlim(0,10) +
  theme_bw()

# FRESH
ggplot(ghg_wsom_extracts, aes(x = delta_do_hr, y = FRESH, color = transect_location)) +
  geom_point() +
  # add limit because of some outliers
  xlim(0,10) +
  theme_bw()

# FI
ggplot(ghg_wsom_extracts, aes(x = delta_do_hr, y = FI, color = transect_location)) +
  geom_point() +
  # add limit because of some outliers
  xlim(0,10) +
  theme_bw()

```

```{r do-indices-boxplots}
# do boxplot
do <- ggplot(filter(ghg_wsom_extracts, type == "dry")) +
  geom_boxplot(aes(x = transect_location, y = delta_do_hr, color = transect_location), lwd = 1, fatten = 1, outlier.shape = NA,
               width = 0.5, show.legend = FALSE) +
  geom_point(aes(x = transect_location, y = delta_do_hr, color = transect_location), position = position_jitterdodge(), size = 2,
             show.legend = FALSE) +
  scale_color_manual(values = palette) +
  # remove outlier
  ylim(0,50) +
  labs(x = "Gradient Location", y = "Dissolved Oxygen Consumption (mg/L/hr)")

# SUVA 254
suva <- ggplot(filter(ghg_wsom_extracts, type == "dry")) +
  geom_boxplot(aes(x = transect_location, y = SUVA254, color = transect_location), lwd = 1, fatten = 1, outlier.shape = NA,
               width = 0.5, show.legend = FALSE) +
  geom_point(aes(x = transect_location, y = SUVA254, color = transect_location), position = position_jitterdodge(), size = 2,
             show.legend = FALSE) +
  scale_color_manual(values = palette) +
  labs(x = "Gradient Location", y = "Specific UV Absorbance at 254 nm")

# Fluorescence Index
FI <- ggplot(filter(ghg_wsom_extracts, type == "dry")) +
  geom_boxplot(aes(x = transect_location, y = FI, color = transect_location), lwd = 1, fatten = 1, outlier.shape = NA,
               width = 0.5, show.legend = FALSE) +
  geom_point(aes(x = transect_location, y = FI, color = transect_location), position = position_jitterdodge(), size = 2,
             show.legend = FALSE) +
  scale_color_manual(values = palette) +
  labs(x = "Gradient Location", y = "Fluorescence Index")

# combine all plots together
ggarrange(do, suva, FI)

```

### PCA with Spectral Indices and GHG/DO Information

```{r calc-pc}
# pull variables of interest for PCA analysis
ghg_wsom_wet_only <- ghg_wsom_extracts %>%
  filter(type == "wet")

pca_ghg_wsom <- ghg_wsom_wet_only %>%
  select(SUVA254:FRESH, delta_do_hr:d_pn2o)

# calculate the principal components
pc <- prcomp(pca_ghg_wsom, scale = TRUE)

# reverse signs of eigenvectors
pc$rotation <- -1*pc$rotation

# look at the principal components
pc$rotation

# reverse signs of each observation's principal components scores
pc$x <- -1*pc$x

# calculate variance represented by each of the principal components
pc$sdev^2 / sum(pc$sdev^2)
```

```{r plot-pca}
pca_plot <- autoplot(pc, data = ghg_wsom_wet_only, colour = "transect_location",
                     frame = T, size = 3, loadings = TRUE,
                     loadings.colour = "black", loadings.label = TRUE,
                     loadings.label.colour = "black",
                     loadings.label.vjust = -0.5,
                     loadings.label.hjust = -0.2) # negative values move right and up

pca_plot + scale_fill_manual(values = palette) +
  scale_color_manual(values = palette) +
  labs(color = "Gradient Location", fill = "Gradient Location")

```



