---
title: "WSOM and GHG Exploration"
output: html_document
date: "2023-03-06"
---

This file reads in spectral data from WSOM extracts and GHG data from lab experiments.
The goal is to look at preliminary relationships between the composition of WSOM across
the transect locations in comparison with metrics related to microbial processing (oxygen consumption, GHG production).

## Set-Up

```{r setup, include=FALSE}
# load packages
require(pacman)
pacman::p_load(tidyverse,
               janitor,
               broom,
               ggpubr, # to use ggarrange for plots
               ggfortify, # for plotting PCA
               GGally, ggcorrplot) # for nice-looking correlation matrices

# load necessary data
ghg <- read_csv("../Downloaded Data/EC1_GHG_data_PR.csv")
wsom_cdom <- read_csv("../WSOM Extracts Data/EC1_WSOC_Extracts_Nov2022_SpectralIndices.csv")
```

```{r}
# set ggplot theme
theme_set(theme_bw())

# set palette
palette <- c("#056009","#0385B5","#021677","#ABABD9","#27220F","#8E7941",
             "#DCCD6A","#8F8D90")

```


## Data Frame Edits

```{r edit-wsom-df}
# pull kit_id and transect location out of cdom sample_id
wsom_cdom$kit_id <- NA
wsom_cdom$transect_location <- NA

for (a in 1:nrow(wsom_cdom)) {
  
  # pull out the second piece of the sample ID split apart at the underscore and use as the plot
  wsom_cdom$kit_id[a] <- strsplit(wsom_cdom$Sample_ID, " ")[[a]][1]

  # pull out the fourth piece of that string and use as the timepoint
  wsom_cdom$transect_location[a] <- strsplit(wsom_cdom$Sample_ID, " ")[[a]][2]
  
}

# mutate transect location abbreviations to match ghg data frame
wsom_cdom_edit <- wsom_cdom %>%
  mutate(transect_location = case_when(
    transect_location == "T" ~ "Transition",
    transect_location == "W" ~ "Wetland",
    transect_location == "UP" ~ "Upland"
  )) %>%
  # remove tea standards
  filter(!is.na(transect_location))

# create unique list of kit IDs and transect locations
unique_extracts <- wsom_cdom_edit %>%
  select(kit_id, transect_location) %>%
  unique()

```

```{r edit-ghg-df}
# filter ghg data to just kits that also have extract data
ghg_filter <- ghg %>%
  filter(kit_id %in% unique_extracts$kit_id) %>%
  filter(transect_location %in% unique_extracts$transect_location)

# trim wsom data frame to match ghg and merge the two together
ghg_wsom_extracts <- wsom_cdom_edit %>%
  filter(kit_id %in% ghg_filter$kit_id) %>%
  # remove the replicate rows - could edit to take the average of these rows to move forward
  filter(!grepl("Rep", Sample_Description)) %>%
  # select specific indices
  select(kit_id, transect_location, Sample_ID:SUVA254, S350_400, FI:FRESH) %>%
  full_join(ghg_filter, by = c("kit_id", "transect_location")) %>%
  # reorder transect location
  mutate(transect_location = factor(transect_location, levels = c("Upland", "Transition", "Wetland")))

# make long data frame for the indices to be able to facet the exploratory plots
#ghg_wsom_long <- ghg_wsom_extracts %>%
  #pivot_longer(cols = c(SUVA254:FRESH), names_to = "spectral_index", values_to = "index_value")

```

## Exploratory Plots

### Dissolved Oxygen Consumption

First thoughts - higher rates of DO consumption should correlate with higher levels of microbial processing.
With more microbial processing, SUVA254 should decline and inversely S350-400 would increase.
FRESH, FI, and HIX should also increase?!?! The question is are these patterns related to transect location as well?

```{r do-vs-indices-scatterplot}
# SUVA254
ggplot(ghg_wsom_extracts, aes(x = delta_do_hr, y = SUVA254, color = transect_location)) +
  geom_point() +
  # add limit because of some outliers
  xlim(0,10) +
  theme_bw()

# FRESH
ggplot(ghg_wsom_extracts, aes(x = delta_do_hr, y = FRESH, color = transect_location)) +
  geom_point() +
  # add limit because of some outliers
  xlim(0,10) +
  theme_bw()

# FI
ggplot(ghg_wsom_extracts, aes(x = delta_do_hr, y = FI, color = transect_location)) +
  geom_point() +
  # add limit because of some outliers
  xlim(0,10) +
  theme_bw()

```

### Boxplots

```{r do-indices-boxplots}
# make a new data frame
new_ghg_wsom <- ghg_wsom_extracts %>%
  filter(type == "dry") %>%
  filter(kit_id != "K019") %>%
  filter(transect_location != "Upland")

# do boxplot
do <- ggplot(new_ghg_wsom) +
  geom_boxplot(aes(x = transect_location, y = delta_do_hr, color = transect_location), lwd = 1, fatten = 1, outlier.shape = NA,
               width = 0.5, show.legend = FALSE) +
  geom_point(aes(x = transect_location, y = delta_do_hr, color = transect_location), position = position_jitterdodge(), size = 2,
             show.legend = FALSE) +
  scale_color_manual(values = c("#0385B5","#021677")) +
  # remove outlier
  ylim(0,50) +
  labs(x = "Gradient Location", y = "Dissolved Oxygen Consumption (mg/L/hr)")

# SUVA 254
suva <- ggplot(new_ghg_wsom) +
  geom_boxplot(aes(x = transect_location, y = SUVA254, color = transect_location), lwd = 1, fatten = 1, outlier.shape = NA,
               width = 0.5, show.legend = FALSE) +
  geom_point(aes(x = transect_location, y = SUVA254, color = transect_location), position = position_jitterdodge(), size = 2,
             show.legend = FALSE) +
  scale_color_manual(values = c("#0385B5","#021677")) +
  labs(x = "Gradient Location", y = "Specific UV Absorbance at 254 nm")

# Fluorescence Index
FI <- ggplot(new_ghg_wsom) +
  geom_boxplot(aes(x = transect_location, y = FI, color = transect_location), lwd = 1, fatten = 1, outlier.shape = NA,
               width = 0.5, show.legend = FALSE) +
  geom_point(aes(x = transect_location, y = FI, color = transect_location), position = position_jitterdodge(), size = 2,
             show.legend = FALSE) +
  scale_color_manual(values = c("#0385B5","#021677")) +
  labs(x = "Gradient Location", y = "Fluorescence Index")

# Change in CO2 Pressure
co2 <- ggplot(new_ghg_wsom) +
  geom_boxplot(aes(x = transect_location, y = d_pco2, color = transect_location), lwd = 1, fatten = 1, outlier.shape = NA,
               width = 0.5, show.legend = FALSE) +
  geom_point(aes(x = transect_location, y = d_pco2, color = transect_location), position = position_jitterdodge(), size = 2,
             show.legend = FALSE) +
  scale_color_manual(values = c("#0385B5","#021677")) +
  labs(x = "Gradient Location", y = "Change in CO2 Pressure")

# Change in CH4 Pressure
ch4 <- ggplot(new_ghg_wsom) +
  geom_boxplot(aes(x = transect_location, y = d_pch4, color = transect_location), lwd = 1, fatten = 1, outlier.shape = NA,
               width = 0.5, show.legend = FALSE) +
  geom_point(aes(x = transect_location, y = d_pch4, color = transect_location), position = position_jitterdodge(), size = 2,
             show.legend = FALSE) +
  scale_color_manual(values = c("#0385B5","#021677")) +
  labs(x = "Gradient Location", y = "Change in CH4 Pressure")

# combine all plots together
ggarrange(do, co2, ch4, suva, FI)

```

### PCA with Spectral Indices and GHG/DO Information

```{r calc-pc}
# pull variables of interest for PCA analysis
ghg_wsom_wet_only <- ghg_wsom_extracts %>%
  filter(type == "wet")

pca_ghg_wsom <- ghg_wsom_wet_only %>%
  select(SUVA254:FRESH, delta_do_hr:d_pn2o)

# calculate the principal components
pc <- prcomp(pca_ghg_wsom, scale = TRUE, retx = TRUE)

# look at the principal components
pc$rotation

pc$x <- -1*pc$x

# calculate variance represented by each of the principal components
pc$sdev^2 / sum(pc$sdev^2)
```

```{r plot-pca}
pca_plot <- autoplot(pc, data = ghg_wsom_wet_only, colour = "transect_location",
                     frame = T, size = 3, loadings = TRUE,
                     loadings.colour = "black", loadings.label = TRUE,
                     loadings.label.colour = "black",
                     loadings.label.vjust = -0.5,
                     loadings.label.hjust = -0.2) # negative values move right and up

pca_plot + scale_fill_manual(values = palette) +
  scale_color_manual(values = palette) +
  labs(color = "Gradient Location", fill = "Gradient Location")

summary(pc)

```

### New PCA

```{r calc-pc}
pca_new_ghg_wsom <- new_ghg_wsom %>%
  select(SUVA254:FRESH, delta_do_hr:d_pn2o)

# calculate the principal components
pc2 <- prcomp(pca_new_ghg_wsom, scale = TRUE, retx = TRUE)

# look at the principal components
pc2$rotation

pc2$x <- -1*pc2$x

# calculate variance represented by each of the principal components
pc2$sdev^2 / sum(pc2$sdev^2)

pca2_plot <- autoplot(pc2, data = new_ghg_wsom, colour = "transect_location",
                     frame = T, size = 3, loadings = TRUE,
                     loadings.colour = "black", loadings.label = TRUE,
                     loadings.label.colour = "black",
                     loadings.label.vjust = -0.5,
                     loadings.label.hjust = -0.2) # negative values move right and up

pca2_plot + scale_fill_manual(values = palette) +
  scale_color_manual(values = palette) +
  labs(color = "Gradient Location", fill = "Gradient Location")
```
### Correlation Matrix

```{r ggpairs, warning = FALSE, message = FALSE}
# set variables to calculate correlations for
ghg_cdom_corr <- pca_ghg_wsom

# assign values of the grouping variables
ghg_cdom_groups <- ghg_wsom_wet_only[,2]

# draw the correlation matrix
ggpairs(ghg_cdom_corr, ggplot2::aes(color = ghg_cdom_groups$transect_location),
        #lower = list(continuous = "smooth"), # adds lm line to the scatterplots
        columnLabels = c("SUVA 254 nm", "Spectral Slope 350-400 nm", "FI", "HIX", "FRESH",
                         "DO Consumption (mg/L/hr)", "Delta CO2 Pressure", "Delta CH4 Pressure", "Delta N2O Pressure")) +
  ggplot2::scale_fill_manual(values = palette) +
  ggplot2::scale_color_manual(values = palette)
```

```{r}
# try a more basic correlation plot
ghg_cdom_corr_matrix <- cor(ghg_cdom_corr)
ghg_cdom_corr_pvalue <- cor_pmat(ghg_cdom_corr)

# basic correlation plot with correlation coefficients
ggcorrplot(ghg_cdom_corr_matrix, hc.order = TRUE,
    type = "lower", lab = TRUE)

# make the correlation plot with significance values
ggcorrplot(ghg_cdom_corr_matrix, hc.order = TRUE,
    type = "lower", method = "circle", p.mat = ghg_cdom_corr_pvalue)
```

### Barplot

```{r}
# select specific measurements to use in the bar plot
sum_ghg_wsom <- new_ghg_wsom %>%
  pivot_longer(cols = c(SUVA254, FRESH, delta_do_hr, d_pco2),
               names_to = "measurement",
               values_to = "value") %>%
  select(kit_id, transect_location, measurement, value) %>%
  group_by(transect_location, measurement) %>%
  summarize(mean_value = mean(value),
            n = n(),
            sd_value = sd(value)) %>%
  mutate(se = sd_value/sqrt(n)) %>%
  mutate(measurement = case_when(
    measurement == "d_pco2" ~ "Delta CO2 Partial Pressure",
    measurement == "SUVA254" ~ "SUVA at 254nm",
    measurement == "FRESH" ~ "Freshness Index",
    TRUE ~ "DO Consumption (mg/L/hr)"
  ))

# plot for change in CO2 partial pressure
d_pco2 <- ggplot(filter(sum_ghg_wsom, measurement == "Delta CO2 Partial Pressure"), aes(x = measurement, y = mean_value, fill = transect_location)) +
  geom_bar(stat = "identity", position = position_dodge(), show.legend = FALSE) +
  geom_errorbar(aes(ymin = mean_value - se, ymax = mean_value + se), color = "grey", 
                width = 0.025, linewidth = 0.8, position = position_dodge(0.9)) + 
  scale_fill_manual(values = palette) +
  labs(fill = "Transect Location", y = "", x = "")

# plot for change in DO
delta_do <- ggplot(filter(sum_ghg_wsom, measurement == "DO Consumption (mg/L/hr)"), aes(x = measurement, y = mean_value, fill = transect_location)) +
  geom_bar(stat = "identity", position = position_dodge(), show.legend = FALSE) +
  geom_errorbar(aes(ymin = mean_value - se, ymax = mean_value + se), color = "grey", 
                width = 0.025, linewidth = 0.8, position = position_dodge(0.9)) + 
  scale_fill_manual(values = palette) +
  labs(fill = "Transect Location", y = "", x = "")

# plot for CDOM indices
wsom <- ggplot(filter(sum_ghg_wsom, measurement == "SUVA at 254nm"|measurement == "Freshness Index"), aes(x = measurement, y = mean_value, fill = transect_location)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  geom_errorbar(aes(ymin = mean_value - se, ymax = mean_value + se), color = "grey", 
                width = 0.025, linewidth = 0.8, position = position_dodge(0.9)) + 
  scale_fill_manual(values = palette) +
  labs(fill = "Transect Location", y = "", x = "")

# arrange plots all together
ggarrange(wsom,
          ggarrange(d_pco2, delta_do, ncol = 2, labels = c("B", "C")),
          nrow = 2,
          labels = "A")

```

